A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-27 analysis/paper/paper_latex.Rmd

In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that Slingshot is the most accurate method to date, but other leading methods (e.g. TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases. We therefore developed a set of guidelines to aid users select the most appropriate method. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be computationally modelled by ordering cells along a trajectory, also called trajectory inference (TI) or pseudotemporal ordering [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`). All methods have their own unique set of characteristics (`r ref("stable", "implementations")`) in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).  A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is the only TI method currently able to deal with disconnected graphs.

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness.. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

@@

In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`, and `r ref("sfig", "extended_evaluation_overview")`). The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`). Only methods published before june 2017 are included in the current version of the evaluation. The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.

In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`). To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets (`r ref("sfig", "synthetic_data_generation")`).

```{r}
add_sfig(
  figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
  "dataset_characterisation",  
  "Characteristics of the real datasets used for the evaluation. Top: distributions of number of cells and genes (after filtering) and the date these datasets were published. Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
)
```

```{r}
add_sfig(
  figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
  "synthetic_data_generation",
  "Workflow to generate the synthetic data. Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated. We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding. Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
)
```

Finally, using both the benchmark results and quality control, we produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.

@@

----

@@

@@

-------

<!-- §1: Overall performance -->
Across all datasets and evaluation metrics, we found that Slingshot predicted the most accurate trajectories, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview", "b")`). When we focused on the benchmark scores per trajectory type, slingshot was the only method that performed well across most trajectory types, followed by TSCAN and Monocle DDRTree. However, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS for linear trajectories, reCAT for cycles, and Monocle DDRTree for trees. This was also reflected in the rankings of different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, performing better at our metric investigating topological similarity, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("sfig", "metric_comparison")`).

<!-- §2: The stability and robustness of the evaluation/method ranking -->
```{r}
correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
filter(n_tasks == round(max(n_tasks)/10)) %>% 
          pull(cor) %>% 
          mean()
top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
filter(method_id == "slngsht") %>% 
filter(n_tasks == round(max(n_tasks)/10)) %>%
pull(rank_match)
random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
filter(method_id == "scoup") %>% 
pull(max_n_tasks_top)

@@

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`. To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
```{r}
add_sfig(
  figure_file(
    "real_synthetic_comparison.tmp.pdf",
    "11-evaluation_robustness"
  ), 
  "real_synthetic_comparison", 
  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
)
```

```{r}
add_sfig(
  figure_file(
    "performance_dataset_variability_combined.tmp.pdf",
    "11-evaluation_robustness"
  ), 
  "performance_dataset_variability_combined", 
  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
)
```

```{r}
add_sfig(
  figure_file('metrics_comparison.tmp.pdf', "5-optimise_parameters/7-metric_comparison"), 
  "metric_comparison", 
  "Comparison of method performance between different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
)
```

### Trajectory types and topology complexity

During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution. If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”. Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).

### Trajectory types and topology complexity

In most cases, the methods which were specifically designed to handle a particular trajectory type, also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`). These methods had typically better edge flip scores - as can be expected - and RF MSE scores, compared to methods not able to handle the particular trajectory type. However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
To further investigate the effect of trajectory complexity on performance, we divided them in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods. While there were no significant differences in performance on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`). Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.

```{r}
add_sfig(
  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
  "trajtype_handle_comparison", 
  "Comparison between the performance of methods able to handle a trajectory type with those who don't. Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
)
```

```{r}
add_sfig(
  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
  "linear_split_comparison", 
  "Comparison of the performance between linear and non-linear methods. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods. Their performance is shown on linear and non-linear datasets according to several performance metrics (right). NS: Not significant, \\*: p-value < 0.05, \\*\\*: p-value < 0.001, \\*\\*\\*: p-value < 0.0001"
)
```

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`). Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets. Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies. Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`). Inferring the correct topology without parameter tuning, is therefore still an open challenge. Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.

@@

----

```{r}
add_sfig(
  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
  "topology_sensitivity", 
  "Sensitivity of each TI method in predicting the correct trajectory topology. Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data. When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
)
```

### Effect of prior information

In the current version of the evaluation, we only provided prior information when a method required it. We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`). Rather, methods which received prior information were on average positioned in the middle of the ranking.  Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.

```{r}
add_sfig(
  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
  "required_priors_performance_comparison", 
  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
)
```

### Algorithm components

We assessed whether the components of an algorithm could  be predictive of a method’s performance, using both random forest classification and statistical testing. Methods which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`). On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.

```{r}
add_sfig(
  figure_file("method_components_scores.tmp.pdf","6-developer_guidelines"), 
  "method_components_scores", 
  "Score of methods containing particular algorithmic components. Shown are the algorithmic components present in four or more methods, and whether the performance of methods containing a certain component was significantly higher than all the other methods. NS: Not significant. *: corrected p-value < 0.05 "
)
```

@@

While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018]. To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review. This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.

We therefore looked at the implementation of each method, and assessed its  quality using a transparent scoring scheme (`r ref("stable", "qc_checks")`). The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, method's behaviour at runtime and the depth by which the method was presented in its study) or which purpose(s) they serve (user friendliness, developer friendliness or future proof). These categorisations can help current developers to improve their tool, and guide the selection of users and developers to use these tools for their purpose. After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.

We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`). We found that recent methods had a slightly better quality than older methods (`r ref("sfig", "qc_over_time")`). However, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

----

```{r}
add_sfig(
  figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
  "qc_over_time", 
  "Quality control scores and number of citations over time."
)
```
We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`). We could also not find any quality aspect which was significantly associated with method performance.

```{r}
add_sfig(
  figure_file(
    "qc_performance_comparison.tmp.pdf",
    "5-optimise_parameters/6-qc_performance_comparison"
  ), 
  "qc_performance_comparison", 
  "Comparison between method quality control scores and overall method performance."
)
```

@@

Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and Slingshot performing equally well on less complex trajectories . No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories. In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS for linear trajectories, and Slingshot for bifurcating trajectories, although Slingshot could return other topologies if it would fit the data more accurately. The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory. Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.

@@

```{r}
add_sfig(
  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
  "implementation_components", 
  "Common algorithmic components shared by different TI methods. Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue). a) Wordcloud of the common components. b) Number of times a component was shared between methods. c) Number of methods containing a particular category."
)
```

@@

```{r}
add_sfig(
  figure_file("modulenets.tmp.pdf", experiment_id = "2-dataset_characterisation/3-synthetic"),
  "modulenets", 
  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
  15, 
  5
)
```

>>>>

In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that slingshot is the most accurate method to date, but other leading methods (e.g. TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases. We therefore developed a set of guidelines to aid users select the most appropriate method. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change. A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation, which compares a large set of methods against a large set of datasets, of TI methods is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

@@

In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`). The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.

@@

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
<!-- Overall evaluation results -->
Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`). When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees. This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).

@@

----------

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).

<!-- Evaluation robustness -->
```{r}
correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
filter(n_tasks == round(max(n_tasks)/10)) %>% 
          pull(cor) %>% 
          mean()
top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
filter(method_id == "slngsht") %>% 
filter(n_tasks == round(max(n_tasks)/10)) %>%
pull(rank_match)
random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
filter(method_id == "scoup") %>% 
pull(max_n_tasks_top)

@@

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`). To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets. The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.

## Relationships between method characteristics to performance

<!-- Inverse relation between edge flip and correlation -->
In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`). These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type. However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories. To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods. While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`). Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.

<!-- Topological complexity versus performance -->
Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`). Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets. Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies. Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases. Inferring the correct topology without parameter tuning is therefore still an open challenge. Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.

@@

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method. Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`). On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.

@@

While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018]. To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_checks")`). This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.

We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`). While recent methods had a slightly better quality than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

@@

We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`). We could also not find any quality aspect which was significantly associated with method performance.

@@

Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and slingshot performing equally well on less complex trajectories . No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")`. should be able to handle graph trajectories or even disconnected trajectories . In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, slingshot and cellTree for linear trajectories, and slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`). The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory. Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.

@@

@@
